M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2012-11-28 notes/thesis-latex-template/thesis.tex
Thesis LaTeX rework

% options are:
% PhD, MSc (choose one)
% beforeExam (to make the personal thanks invisible)
\documentclass[MSc,beforeExam]{iitcsthesis}

\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{units}   
\usepackage{float}   
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage[colorlinks]{hyperref}
\usepackage{framed}
\usepackage{url}
\usepackage{longtable}

@@

% The English abstract should be 200-500 words long.
\abstractEnglish
In this work we study the electrokinetic 
migration of particles in an electrolyte solution 
due to the application of an external electric field,
and propose a numerical framework for iterative solution of such problems.
The electrokinetic transport process can be used for
transporting and manipulating micro- and nanoscale objects in 
many nanotechnology applications, nano-fluidic devices, packed bed separation, 
desalination processes and various electrophoresis applications.
Due to strong electrostatic forces, a thin boundary layer forms near the particle
surface. This results in scale disparity of the boundary layer, which hinders full numerical solution. 
We employ nonlinear macroscale effective boundary conditions that are derived
using the specific chemical properties of the particle
(for an ion-exchanger particle at \cite{yariv2010migration}
and for a surface charged inert particle at \cite{schnitzer2012surface}).
The resulting macroscale nonlinear partial differential system 
is discretized and an iterative Newton solver is
constructed automatically from the discrete equations.
Numerical results are obtained for an ion-exchanger and 
for a surface charged inert particle. 
An asymptotic analytical solution is 
used for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, 
and good correspondence is achieved.

@@

The variables of the electrokinetic flow equations are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle size. 
An effective boundary condition outside the Debye layer
has been developed for spherical particles, yielding a macroscale 
model at \cite{yariv2010migration, schnitzer2012surface}.
The model is solved for weak electric field in an axisymmetric setting 
but it is hard to extend this analytic solution to more general systems.
Once the electric field becomes stronger, 
significant nonlinear phenomena emerge.
This interesting regime has not yet been explored extensively.

@@

The manipulation can be achieved by either applying the external field at inlets and outlets, or it can be applied locally in the microchannel by integrated components.
A typical microchannel is produced with lithographic techniques. Normally the Reynolds
number is very low (on the order of unity or smaller) which means that viscosity dominates
and the convective term of the Navier–Stokes equations are insignificant. This means that
microflows most often are laminar.

In the last decade there has been an increase in microfluidics research, 
due to the spread of tools for microfluidic systems fabrication, 
production of cheap and portable devices for fast point-of-care analysis, 
and fundamental research of physical, chemical and biological processes.
Various applications are \cite{erickson2004integrated, squires2005microfluidics, hardt2007microfluidic}:
\begin{enumerate}
\item DNA and cell analysis \cite{horsman2007forensic, wu2004chemical}.
\item Separation-based detection \cite{weinberger1993practical}.
\item Microscale chemical processing \cite{kothare2006dynamics}.
\item Packed bed separation \cite{leinweber2004, losey2001microfabricated}.
\item Desalination processes \cite{kim2010direct}.
\item Drug delivery \cite{santini2000microchips}.
\end{enumerate}

@@

\subsection{Nondimensionalization}
Spatial coordinates $\br$ are normalized by $a^*$ 
(the characteristic size of the particle).
Positive and negative ionic concentrations are denoted by $c_+$ and $c_-$, respectively, and
are normalized by the ambient ionic concentration $c^*$. The ionic valences are $\pm z$.
The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$, 
where $D^*$ is the ionic diffusivity):
\begin{equation}
\bj^*_\pm = 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
the fluid remains electroneutral ($q \approx 0$, thus $c_+ \approx c_-$), 
except inside a thin boundary layer of width $O(\delta)$ 
surrounding the particle (the ``Debye layer'').
This boundary layer makes it very hard to construct 
a direct numerical solver for the full electrokinetic problem.
However, we exploit the scale disparity at the Debye layer, 
which allows the PDE system to be rewritten in the ``inner'' region 
as an ODE system using the asymptotic limit of $\delta \rightarrow 0$.
This system can be integrated and solved analytically,
yielding ``effective'' boundary conditions near
the particle surface for the ``outer'' bulk region equations.

@@

\section{Work Outline}

@@

The work is structured as follows: 
The numerical scheme for the discretization and 
the solution of the system is described in Chapter \ref{ch:algorithm}.
Numerical results and their comparison to the analytical asymptotic solutions 
are presented and discussed in Chapter \ref{ch:results}.

@@

We start by presenting the numerical framework used for the nonlinear solver construction.

@@

Finite Volume Methods \cite{ferziger1999computational}
 are based on a discretization of the integral forms of the conservation equations.
The simulation domain is divided into a set of discrete non-overlapping control volumes.
Each conservartion equation is discretized over the control volumes,
such that the integral form of the conservation law is satisfied using an approximation of the fluxes
over the control volume faces.
This method ensures that the discretization is conservative in the discrete sense.

\subsection{Ghost Points}

\subsection{Upwind Discretization}

\subsection{Newton's Method}
Newton's method is used to solve a nonlinear system of equations of the form $\cO(\bx) = \bzero$.
For small step $\Delta\bx$, we can linearize the operator by
\begin{align}
\cO(\bx + \Delta\bx) \approx \cO(\bx) + \bnabla \cO(\bx) \Delta\bx
\end{align}
Thus, we use the following iteration, starting from an initial solution $\bx_0$:
\begin{align}
\Delta\bx_n &= - \left[\bnabla \cO(\bx_n)\right]^{-1} \cdot \cO(\bx_n) \\
\bx_{n+1} &= \bx_n + \Delta\bx_n
\end{align}

Then, a sparse linear system $\boldsymbol{A} \Delta \bx = \boldsymbol{b}$
needs to be solved to yield the update $\Delta \bx$.
When the method converges, $\bx_n \rightarrow \bx_\infty$, 
the asymptotic convergence is in general quadratic:
\begin{equation}
\|\bx_{n+1} - \bx_\infty\| \le k \|\bx_{n} - \bx_\infty\|^2.
\end{equation}
Thus, when the initial guess is close enough to the solution, the solver
will typically converge in very few steps.

\subsection{Richardson Extrapolation}

@@

\subsection{Spherical Coordinates}
Because the system is axisymmetric, we employ
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{align}
\bnabla f &= \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, \\
\bnabla \cdot \bF &= \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{align}
The scalar Laplacian can be written as:
\begin{align}
\Laplacian f = \bnabla \cdot (\bnabla f) = 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{align}
The vector gradient and vector Laplacian can be written as:
\begin{align}
\nonumber
\bnabla \bF &= \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat, 
\\
\brhat \cdot \bLaplacian \bF &=
\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}
\\
\nonumber
\bthetahat \cdot \bLaplacian \bF &= 
\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}.
\end{align}

Note that in the spherical coordinate system,
the vector Laplacian operator cannot be computed in a decoupled manner as in the Cartesian system,
since the unit vectors $\brhat, \bthetahat, \bphihat$ are functions of the coordinates
in a curvilinear coordinate system.
See Appendix \ref{append:spherical} for the full derivation of the identities above.

@@

\begin{figure}[htbp]
    \begin{center}
	\includegraphics[width=0.8\textwidth]{figs/grids.eps}
        \caption[Computational grid]{The computational grid is defined using
        spherical coordinates system, due to axisymmetry of the physical problem
        and boundary conditions discretization on the particle surface.}
    \label{fig:grids}
    \end{center}
\end{figure}

@@

\subsection{System Equations} \label{sec:disc_equations}
Charge and salt fluxes \eqref{eq:bulk_flux} are discretized on grid cell boundaries 
(using an upwind scheme $\mathcal{U}$ for numerical stability at large cell Peclet number)  
using \eqref{eq:disc_diff} and \eqref{eq:disc_interp}:
\begin{align} 
\nonumber
I^h_r = -\cI_r(C^h) \cdot \cD_r(\varPhi^h), \\
\nonumber
I^h_\theta = -\cI_\theta(C^h) \cdot \frac{\cD_\theta(\varPhi^h)}{r}, \\
\nonumber
J^h_r = -\cD_r(C^h) + \alpha V^h_r \cdot \mathcal{U}^{\bV^h}_r (C^h), \\
J^h_\theta = -\frac{\cD_\theta(C^h)}{r} + \alpha V^h_\theta \cdot \mathcal{U}^{\bV^h}_\theta (C^h), 
\\ \nonumber
 \mathcal{U}^{\bV^h}_r(C^h)[i,j] = C^h\left[i-\frac{\sign(V^h_r)}{2}, j\right], 
\\ \nonumber
 \mathcal{U}^{\bV^h}_\theta(C^h)[i,j] = C^h\left[i, j-\frac{\sign(V^h_\theta)}{2}\right]. 
\end{align}

Mass flux $\bV$ is discretized on grid cells boundaries, 
using the staggered velocity grid.

Force components ($\bF = -\bnabla P + \bLaplacian \bV + \bnabla \varPhi \Laplacian \varPhi$) 
are discretized on the staggered velocity grid using finite difference
method, where linear interpolation is used for the Coulomb force and 
for vector Laplacian components.
\begin{align}
\nonumber
F^h_r &= -\cD_r(P^h) 
          + \cL(V^h_r) - \frac{2}{r^2} V^h_r 
		  - \frac{2}{r^2 \sin\theta} \cI_r(\cD_\theta (V^h_\theta \sin\theta))
          + \cD_r(\varPhi^h) \cdot \cI_r(\cL(\varPhi^h)), \\
\nonumber
F^h_\theta &= -\frac{\cD_\theta(P^h)}{r} 
		  + \cL(V^h_\theta) - \frac{F^h_\theta}{r^2 \sin^2\theta} 
		  + \frac{2}{r^2} \cI_\theta(\cD_\theta(F^h_r))
		  + \frac{\cD_\theta(\varPhi^h)}{r} \cdot \cI_\theta(\cL(\varPhi^h)), \\
\cL(f^h) &= \frac{1}{r^2}\cD_r\pars{\cD_r(f^h) r^2} + 
\frac{1}{r^2 \sin\theta} \cD_\theta\pars{\cD_\theta(f^h) \cdot \sin\theta}.
\end{align}

\subsection{Boundary Conditions} \label{sec:disc_boundary}
In order to discretize boundary condtions for $\varPhi$, $C$ and $\bV$, 
``ghost'' points are employed. 
The grid is extended to include points outside the domain interior,
and the variables at the ``ghost'' points are set to satisfy 
the discretized boundary conditions, following the analysis below.
Note that $P$ has no boundary conditions -- so the pressure variable is defined 
only in the domain interior.

Far away from the particle boundary \eqref{eq:bndcond_inf}, we have:
\begin{equation}
\begin{array}{cccc}
\cD_r \pars{\varPhi^h} = -\beta \cos\bar{\theta}, &
\cI_r \pars{C^h} = 1 &
V_r^h = -\cU \cos\theta, &
V_\theta^h = \cU \sin\theta.
\end{array}\end{equation}

For $\theta = 0$ and $\theta = \pi$, the ``ghost'' points are defined by axial 
symmetry considerations:
\begin{equation} 
\begin{array}{cccc}
\cD_\theta \pars{\varPhi^h} = 0, &
\cD_\theta \pars{C^h} = 0, &
\cD_\theta \pars{V_r^h} = 0, &
V_\theta^h = 0
\end{array}
\end{equation}

The specific boundary conditions at $r=1$
depend on the chemical properties of particle surface.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solver Design}
An operator $\cO$ is defined by its input vector $\bx$ and its output 
$\by = \cO(\bx)$.
In order to solve the equation $\cO(\bx) = \bzero$, Newton's method is applied, 
given an initial solution $\bx_0$.
Each step requires the computation of the residual vector, $\br = \cO(\bx)$, and
the gradient matrix of the operator, $\bG = \bnabla \cO(\bx)$, 
given the current solution $\bx$.

@@

Due to symmetry considerations, the total force $\bF$ is 
aligned with $\ui$ and it vanishes iff $\ui \cdot \bF = 0$.
\begin{eqnarray} \nonumber
F_\imath = \ui \cdot \bF &=& 
\oint_\mathcal{S} \pars{\ui \cdot \tT \cdot \bnhat} dA = 
\int_0^\pi f_\imath(\theta) 2\pi \sin\theta d\theta ,
\\  
f_\imath = \ui \cdot \tT \cdot \brhat &=& \pars{-P + 2\cD_r(V_r) + 
\frac{1}{2}\pars{\cD_r(\varPhi)}^2 - \frac{1}{2r^2}\pars{\cD_\theta(\varPhi)}^2}\cos\theta 
\\  \nonumber
&& -\pars{\cD_r(V_\theta) - \frac{V_\theta}{r}
+ \frac{1}{r}\cD_r(\varPhi) \cD_{\theta}(\varPhi)}\sin\theta.
\end{eqnarray}

The integral above is approximated by 1D numerical quadrature
(using the mid-point rule)
\begin{equation}
F_\imath^h(\beta, \cU^h) = \sum_{i=1}^{n_\theta} f^h(\bar\theta_i) \cdot 
              2 \pi \sin\bar\theta_i \cdot \Delta\theta_i,
\end{equation}
to yield the total force $F_\imath^h(\beta, \cU^h)$ as a function of the drift velocity $\cU^h$.
Since the goal is to find the steady-state solution, $F_\imath^h$ 
is required to be zero --
and the appropriate $\cU^h$ is found by a simple 1D root-finding algorithm,
applied as an outer loop:
\begin{equation} \label{eq:disc_zero_force}
F_\imath^h(\beta, \cU^h(\beta)) = 0
\end{equation}

@@

\subsection{Operator Interface}
The base \verb|Operator| interface is defined by 
supporting the computation of a residual vector \verb|Operator::res()|
and a gradient matrix \verb|Operator::grad()|, given an input vector $\bx$.
This interface is implemented by specific operators classes,
which are used to construct the system operator $\cO$, so that 
the Newton solver (that requires computing the residual and the gradient) 
can be applied automatically.
Each such operator may have other operators as its inputs, so the residual
and the gradient are computed recursively, using the chain rule (applied to operators).

@@

>>>>

% PhD, MSc (choose one)
% beforeExam (to make the personal thanks invisible)
\documentclass[MSc,beforeExam]{iitcsthesis}

\usepackage{graphicx}
\usepackage{tikz}
\usepackage{amsmath} 
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{units}   
\usepackage{float}   
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage[colorlinks]{hyperref}
\usepackage{framed}
\usepackage{url}
\usepackage{longtable}
\usepackage[framed,numbered,autolinebreaks,useliterate]{mcode}

@@

% The English abstract should be 200-500 words long.
\abstractEnglish
In this work we study the electrokinetic 
migration of particles in an electrolyte solution 
due to the application of an external electric field,
and propose a numerical framework for iterative solution of such problems.
The electrokinetic transport process can be used for
transporting and manipulating micro- and nanoscale objects in 
many nanotechnology applications, nano-fluidic devices, packed bed separation, 
desalination processes and various electrophoresis applications.
Due to strong electrostatic forces, a thin boundary layer forms near the particle
surface. This results in scale disparity of the boundary layer, which makes a 
full numerical solution challenging. 
We employ nonlinear macroscale effective boundary conditions that have been derived
using the specific chemical properties of the particle
(for an ion-exchanger particle by Yariv in \cite{yariv2010migration}
and for a surface charged inert particle by Schnitzer and Yariv in \cite{schnitzer2012surface}).
The resulting macroscale nonlinear partial differential system 
is discretized and an iterative Newton solver is
constructed automatically from the discrete equations.
Numerical results are obtained for an ion-exchanger and 
for a surface charged inert particle. 
An asymptotic analytical solution is 
used for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, 
and good correspondence is achieved.

@@

The variables of the electrokinetic flow equations are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle size. 
An effective boundary condition outside the Debye layer
has been developed for spherical particles, yielding a macroscale 
model in \cite{yariv2010migration, schnitzer2012surface}.
The model is solved for weak electric field in an axisymmetric setting 
but it is hard to extend this analytic solution to more general systems.
Once the electric field becomes stronger, 
significant nonlinear phenomena emerge.
This interesting regime has not yet been explored extensively.

@@

The manipulation can be achieved by either applying the external field at inlets and outlets, or it can be applied locally in the microchannel by integrated components.
A typical microchannel is produced with lithographic techniques. Normally the Reynolds
number is very low (on the order of unity or smaller) which means that viscosity dominates
and the convective term of the Navier-Stokes equations are insignificant. This also means that
microflows most often are laminar.

In the last decade there has been an increase in microfluidics research, 
due to the spread of tools for microfluidic systems fabrication, 
production of cheap and portable devices for fast medical analysis, 
and fundamental research of physical, chemical and biological processes.
Various applications are \cite{whitesides2006origins, erickson2004integrated, squires2005microfluidics, hardt2007microfluidic}:
\begin{enumerate}
\item DNA separation and sequencing \cite{wainright2003preconcentration}.
\item Cell analysis and forensic identification \cite{wu2004chemical, weinberger1993practical, horsman2007forensic}.
\item Integrated microchemical systems \cite{kothare2006dynamics}.
\item Packed-bed separation \cite{leinweber2004, losey2001microfabricated}.
\item Desalination processes \cite{kim2010direct}.
\item Drug delivery and pathogen detection \cite{santini2000microchips}.
\end{enumerate}

@@

\subsection{Nondimensionalization}
We follow the derivation presented by Yariv in \cite{yariv2010asymptotic}.
Spatial coordinates $\br$ are normalized by $a^*$ 
(the characteristic size of the particle).
Positive and negative ionic concentrations are denoted by $c_+$ and $c_-$, respectively, and
are normalized by the ambient ionic concentration $c^*$. The ionic valences are $\pm z$.
The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$, 
where $D^*$ is the ionic diffusivity):
\begin{equation}
\bj^*_\pm = 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
the fluid remains electroneutral ($q \approx 0$, thus $c_+ \approx c_-$), 
except inside a thin boundary layer of width $O(\delta)$ 
surrounding the particle (the ``Debye layer'').
This boundary layer makes it very hard to construct 
a direct numerical solver for the full electrokinetic problem.
However, the scale disparity at the Debye layer can be exploited
(as shown by Yariv in \cite{yariv2010asymptotic}), 
which allows the PDE system to be rewritten in the ``inner'' region 
as an ODE system using the asymptotic limit of $\delta \rightarrow 0$.
This system can be integrated and solved analytically,
yielding ``effective'' boundary conditions near
the particle surface for the ``outer'' bulk region equations.

@@

\section{Objective of this work}

@@

@@

We start by presenting the numerical framework
that is used for the construction of the nonlinear solver.

@@

Finite Volume methods \cite{ferziger2002computational} 
are used to define a discretization of the conservation equations' integral forms.
The simulation domain is divided into a set of discrete non-overlapping control volumes.
Each conservation equation is discretized over the control volumes,
such that the integral form of the conservation law is satisfied using an approximation of the fluxes
over the control volume faces.
This method ensures that the discretization is conservative in the discrete sense.

\subsection{Flux Discretization}

In order to define the fluxes on the boundaries of each control volume, we employ the 
conservative formulation of the governing equations above. 
The flux terms are defined using finite differences and linear interpolation of the disrete 
variables (defined at the centroid of each cell), except for the advection flux which should not be 
discretized using linear interpolation -- due to a numerical instability 
for large cell Peclet numbers \cite{strikwerda2004finite}.

\subsection{Ghost Points}
In order to discretize boundary condtions for $\varPhi$, $C$ and $\bV$, 
``ghost'' points are employed. 
The grid is extended to include points outside the domain interior,
and the variables at the ``ghost'' points are set to satisfy 
the discretized boundary conditions on the grid boundaries.
Note that $P$ has no boundary conditions -- so the pressure variable is defined 
only in the domain interior.

\subsection{Newton's Method}
Newton's method is used to solve a nonlinear system of equations of the form $\cO(\bx) = \bzero$.
For small step $\Delta\bx$, we can linearize the operator by
\begin{align}
\cO(\bx + \Delta\bx) \approx \cO(\bx) + \bnabla \cO(\bx) \Delta\bx
\end{align}
Thus, we use the following iteration, starting from an initial solution $\bx_0$:
\begin{align}
\Delta\bx_n &= - \left[\bnabla \cO(\bx_n)\right]^{-1} \cdot \cO(\bx_n) \\
\bx_{n+1} &= \bx_n + \Delta\bx_n
\end{align}
Then, a sparse linear system $\boldsymbol{A} \Delta \bx = \boldsymbol{b}$
needs to be solved to yield the update $\Delta \bx$.
When the method converges, $\bx_n \rightarrow \bx_\infty$, 
the asymptotic convergence is in general quadratic:
\begin{equation}
\|\bx_{n+1} - \bx_\infty\| \le k \|\bx_{n} - \bx_\infty\|^2.
\end{equation}
Thus, when the initial guess is close enough to the solution, the solver
will typically converge in very few steps.

\subsection{Richardson Extrapolation}
In order to achieve higher numerical accuracy, we employ Richardson extrapolation \cite{sidi2003practical}
on the resulting steady-state velocity $\cU(h)$, as a scalar function of the grid spacing $h$.
This method enables us to detect and analyze nonlinear effects for $\beta \ll 1$, where 
a higher-order discretization is required because otherwise
the numerical error becomes comparable to the difference between the linear analytical regime 
and the numerical results.

@@

\subsection{Spherical Coordinates}
Because the system is axisymmetric and the particle is a sphere, we employ
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{align}
\bnabla f &= \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, \\
\bnabla \cdot \bF &= \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{align}
The scalar Laplacian can be written as:
\begin{align}
\Laplacian f = \bnabla \cdot (\bnabla f) = 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{align}
The vector gradient and vector Laplacian can be written as:
\begin{align}
\nonumber
\bnabla \bF &= \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat, 
\\
\brhat \cdot \bLaplacian \bF &=
\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}
\\
\nonumber
\bthetahat \cdot \bLaplacian \bF &= 
\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}.
\end{align}
Note that in the spherical coordinate system,
the vector Laplacian operator cannot be computed in a decoupled manner as in the Cartesian system,
because the unit vectors $\brhat, \bthetahat, \bphihat$ are functions of the coordinates
in a curvilinear coordinate system.
See Appendix \ref{append:spherical} for the full derivation of the identities above.

@@

\begin{figure}[htbp]
    \begin{center}
	\includegraphics[width=0.8\textwidth]{figs/grids.eps}
        \caption[Computational grid]{The computational grid is defined using
        a spherical coordinate system, due to axisymmetry of the physical problem
        and boundary conditions discretization on the particle surface.}
    \label{fig:grids}
    \end{center}
\end{figure}

@@

\subsection{System Equations} \label{sec:disc_equations}
Charge and salt fluxes \eqref{eq:bulk_flux} are discretized on grid cell boundaries 
(using an upwind scheme $\mathcal{U}$ for numerical stability at large cell Peclet number)  
using \eqref{eq:disc_diff} and \eqref{eq:disc_interp}:
\begin{align} 
\nonumber
I^h_r = -\cI_r(C^h) \cdot \cD_r(\varPhi^h), \\
\nonumber
I^h_\theta = -\cI_\theta(C^h) \cdot \frac{\cD_\theta(\varPhi^h)}{r}, \\
\nonumber
J^h_r = -\cD_r(C^h) + \alpha V^h_r \cdot \mathcal{U}^{\bV^h}_r (C^h), \\
J^h_\theta = -\frac{\cD_\theta(C^h)}{r} + \alpha V^h_\theta \cdot \mathcal{U}^{\bV^h}_\theta (C^h), 
\\ \nonumber
 \mathcal{U}^{\bV^h}_r(C^h)[i,j] = C^h\left[i-\frac{\sign(V^h_r)}{2}, j\right], 
\\ \nonumber
 \mathcal{U}^{\bV^h}_\theta(C^h)[i,j] = C^h\left[i, j-\frac{\sign(V^h_\theta)}{2}\right]. 
\end{align}
Mass flux $\bV$ is discretized on grid cells boundaries, 
using the staggered velocity grid.
Force components ($\bF = -\bnabla P + \bLaplacian \bV + \bnabla \varPhi \Laplacian \varPhi$) 
are discretized on the staggered velocity grid using finite difference
method, where linear interpolation is used for the Coulomb force and 
for vector Laplacian components.
\begin{align}
\nonumber
F^h_r &= -\cD_r(P^h) 
          + \cL(V^h_r) - \frac{2}{r^2} V^h_r 
		  - \frac{2}{r^2 \sin\theta} \cI_r(\cD_\theta (V^h_\theta \sin\theta))
          + \cD_r(\varPhi^h) \cdot \cI_r(\cL(\varPhi^h)), \\
\nonumber
F^h_\theta &= -\frac{\cD_\theta(P^h)}{r} 
		  + \cL(V^h_\theta) - \frac{F^h_\theta}{r^2 \sin^2\theta} 
		  + \frac{2}{r^2} \cI_\theta(\cD_\theta(F^h_r))
		  + \frac{\cD_\theta(\varPhi^h)}{r} \cdot \cI_\theta(\cL(\varPhi^h)), \\
\cL(f^h) &= \frac{1}{r^2}\cD_r\pars{\cD_r(f^h) r^2} + 
\frac{1}{r^2 \sin\theta} \cD_\theta\pars{\cD_\theta(f^h) \cdot \sin\theta}.
\end{align}

\subsection{Boundary Conditions} \label{sec:disc_boundary}

Far away from the particle boundary \eqref{eq:bndcond_inf}, we have:
\begin{equation}
\begin{array}{cccc}
\cD_r \pars{\varPhi^h} = -\beta \cos\bar{\theta}, &
\cI_r \pars{C^h} = 1 &
V_r^h = -\cU \cos\theta, &
V_\theta^h = \cU \sin\theta.
\end{array}\end{equation}
For $\theta = 0$ and $\theta = \pi$, the ``ghost'' points are defined by axial 
symmetry considerations:
\begin{equation} 
\begin{array}{cccc}
\cD_\theta \pars{\varPhi^h} = 0, &
\cD_\theta \pars{C^h} = 0, &
\cD_\theta \pars{V_r^h} = 0, &
V_\theta^h = 0
\end{array}
\end{equation}
The specific boundary conditions at $r=1$
depend on the chemical properties of particle surface.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solver Design}
Let $\cO : \R^N \rightarrow \R^N $ be an nonlinear operator.
In order to solve the equation $\cO(\bx) = \bzero$, Newton's method is applied, 
given an initial solution $\bx_0$.
Each step requires the computation of the residual vector, $\br = \cO(\bx)$, and
the gradient matrix of the operator, $\bG = \bnabla \cO(\bx)$, 
given the current solution $\bx$.

@@

Due to symmetry considerations, the total force $\bF$ is 
aligned with $\ui$ and it vanishes iff $\ui \cdot \bF = 0$.
\begin{eqnarray} \nonumber
F_\imath = \ui \cdot \bF &=& 
\oint_\mathcal{S} \pars{\ui \cdot \tT \cdot \bnhat} dA = 
\int_0^\pi f_\imath(\theta) 2\pi \sin\theta d\theta ,
\\  
f_\imath = \ui \cdot \tT \cdot \brhat &=& \pars{-P + 2\cD_r(V_r) + 
\frac{1}{2}\pars{\cD_r(\varPhi)}^2 - \frac{1}{2r^2}\pars{\cD_\theta(\varPhi)}^2}\cos\theta 
\\  \nonumber
&& -\pars{\cD_r(V_\theta) - \frac{V_\theta}{r}
+ \frac{1}{r}\cD_r(\varPhi) \cD_{\theta}(\varPhi)}\sin\theta.
\end{eqnarray}
The integral above is approximated by 1D numerical quadrature
(using the mid-point rule)
\begin{equation}
F_\imath^h(\beta, \cU^h) = \sum_{i=1}^{n_\theta} f^h(\bar\theta_i) \cdot 
              2 \pi \sin\bar\theta_i \cdot \Delta\theta_i,
\end{equation}
to yield the total force $F_\imath^h(\beta, \cU^h)$ as a function of the drift velocity $\cU^h$.
Since the goal is to find the steady-state solution, $F_\imath^h$ 
is required to be zero --
and the appropriate $\cU^h$ is found by a simple 1D root-finding algorithm,
applied as an outer loop:
\begin{equation} \label{eq:disc_zero_force}
F_\imath^h(\beta, \cU^h(\beta)) = 0
\end{equation}

@@

\subsection{Operator Interface}
The base \verb|Operator| interface is defined by 
supporting the computation of a residual vector \verb|Operator::res()|
and a gradient matrix \verb|Operator::grad()|, given an input vector $\bx$.
This interface is implemented by specific operator classes,
which are used to construct the system operator $\cO$, so that 
the Newton solver (that requires computing the residual and the gradient) 
can be applied automatically.
Each such operator may have other operators as its inputs, so the residual
and the gradient are computed recursively, using the chain rule (applied to operators).

@@

\subsection{Sample source code}

\subsubsection{Variable class}
\begin{figure}
\begin{lstlisting}
classdef Variable < handle
properties
    value;
end

methods
    function self = Variable(varargin)
        self.value = col(varargin{:});
    end

    function r = res(self)
        r = self.value(:);
    end

    function G = grad(self) 
        G = speye(numel(self.value));
    end

    function update(self, delta)
        assert(numel(self.value) == numel(delta))
        self.value = self.value + delta;
    end
end

end
\end{lstlisting}
\end{figure}

\subsubsection{Operator interface}
\begin{lstlisting}
% An operator base class.
classdef Operator < handle
properties
    grid; % a grid object for the output
end
methods
    function op = Operator(grid)
        op.grid = grid;
    end
    function op = plus(op1, op2)
        op = Sum(op1, op2);
    end
    function op = minus(op1, op2)
        op = Minus(op1, op2);
    end
    function op = uminus(op)
        op = Product(-1, op);
    end
    function op = mtimes(op1, op2)
        op = Product(op1, op2);
    end
    function op = log(op)
        op = Func(op, 'log(x)');
    end
    function op = exp(op)
        op = Func(op, 'exp(x)');
    end
    function op = sinh(op)
        op = Func(op, 'sinh(x)');
    end
    function op = cosh(op)
        op = Func(op, 'cosh(x)');
    end
    function op = tanh(op)
        op = Func(op, 'tanh(x)');
    end
end
end
\end{lstlisting}

\subsubsection{Linear operators}
\begin{lstlisting}
% Linear Sparse operator.
classdef Linear < Operator
properties
    L;
    op;
end
